### 2017-03-20 Manuscripts/cog-psych/cogpsych.tex
expanded/some reformulation of intro and some of background. restructuring

@@

The creation of novel concepts and ideas is a highly intriguing -- yet infrequently studied -- topic of research in human cognition. The creative use of conceptual knowledge is, for example, a core element of scientific investigation, wherein the generation of new ideas is critical to designing experiments and explaining observations. However, due to its complexity, little is presently known about the cognitive processes underlying the generation of new concepts.

Much of what we know about concept generation comes from the foundational literature on creative cognition. In a classic series of reports, Ward \& colleagues \citep{ward1995s,ward1994structured,marsh1999inadvertent,ward2002role,smith1993constraining} established that category generation is highly constrained by prior knowledge: generated categories tend to consist of features observed in known categories, and they tend to exhibit the distributional properties as found in known categories. In a classic study, \cite{ward1994structured} asked participants to generate new species of alien animals by drawing and describing members of the species. People tended to generate species with the same features as on Earth (e.g., eyes, legs, wings), and possessing the same feature correlations as on Earth (e.g., feathers co-occur with wings). Likewise, aliens drawn from the same species tended to share more features with one another compared to members of opposite species.

@@

\cite{jern2013probabilistic} additionally tested a "copy-and-tweak" model that broadly resembles the earlier "Path of Least Resistance" account: the core proposal is that participants generate new items by copying stored examples from memory and tweaking them to generate something new. The copy-and-tweak model differs from the path of least resistance account in that it notably omits the hierarchical organization of categories, as well as selectivity in which features are changed. Instead, their copy-and-tweak model corresponds to a direct exemplar-similarity approach \citep[e.g.,][]{nosofsky1984choice}: participants are thought to generate new items according to their similarity to known members of the target category. This model provided a poor account of the observed data: indeed, the experiments devised by \citeauthor{jern2013probabilistic} were specifically designed to challenge this model. However, it's application is notable as a first step toward explaining category generation using well-known formal approaches from the categorization literature.

@@

It is worth noting that the literature reviewed above provides a somewhat limited view of category generation: the main concern of the published experiments has been on the distributional correspondences between learned and generated categories, and as a result most of the modeling efforts have been towards explaining those effects. In this paper, we investigate another important constraint on the creative use of conceptual knowledge: category contrast. In order to generate a novel concept, individuals must produce something that is in some capacity \textit{different} from what they already know. Thus, in a trivial sense, contrast can be viewed as a fundamental constraint on creative generation: new concepts must firstly be different from existing ones.

Although it is evident that people are \textit{capable} of creating new concepts and categories, it is not entirely clear how new concepts are systematically made different from what is already known. The hierarchical sampling model developed by \cite{jern2013probabilistic}, for example, does not provide any strong claims about how generated categories should contrast with learned ones: the model only proposes that generated categories are sampled from the same underlying domain distribution as observed categories, and will thus share a common distributional structure. The authors do not make predictions about the \textit{location} of the category within the domain (the perceptual instantiation of category members). Indeed, under a strict interpretation of the hierarchical Bayesian approach, the generated category should actually overlap substantially with the known categories (as it's location is drawn based on those categories). Alternatively, the model's placement of the new category may simply be random (this was the case in the simulations conducted by \citeauthor{jern2013probabilistic}).

@@

@@

\nbcnote{Maybe mention something about path of least resistance prediction that contrast is on non-characteristic dimensions? }

\subsection{Overview}

As noted in the sections above, research on the creative generation of novel concepts has thusfar focused on the finding that generated categories tend to possess distributional commonalities with known categories. However, a fundamental goal of concept generation is to create something \textit{new} (i.e., different from what is already known). The manner in which generated categories differ from known ones is, nonetheless, poorly understood: existing formal approaches do not make strong predictions about how creatively generated concepts should systematically differ from existing ones.

@@

@@

\section{The PACKER Model}

The PACKER model is an extension of the influential Generalized Context Model of category learning \citep[GCM;][]{nosofsky1984choice,nosofsky1986attention}. The GCM simulates categorization under the assumption that learners represent categories as a collection of exemplars, corresponding to the labeled stimuli they have observed. The exemplars are encoded within a $k$-dimensional psychological space, and categorization is based on the amount of similarity between the item to be categorized and the stored exemplars. Similarity between two examples, $s\left(x_i, x_j\right)$, is computed as an inverse exponential function of distance \citep[following][]{shepard1957stimulus,shepard1987toward}:

>>>>

\jlanote{This is awesome!!!}

@@

Creativity, innovation, imagination, and the creation of new ideas are among the most fascinating, important and difficult human capabilities to study. They are fascinating and important capabilities because they have produced some of the greatest scientific breakthroughs that revolutionized the world. Unfortunately, scientific breakthroughs are rare and the result of a great deal of cognitive effort, which it difficult to investigate the underlying cognitive representations and processes responsible for them using standard experimental methodology or formalize in computational models. Although we tend to focus on the most salient products of these processes (e.g., scientific breakthroughs), every person is likely to have generated a novel sentence, thought, and/or drawing.\footnote{\jlanote{We need citations/examples for this. In a compositional system, it's pretty clear this has to be the case, but I'd like to use someone else as a reference rather than arguing this myself.}} This observation provides a fantastic opportunity: By examining how people create new concepts in a domain amenable to formalization in computational models, we can investigate (some aspects of) creativity, innovation, and imagination scientifically.

%The creation of novel concepts and ideas is a highly intriguing -- yet infrequently studied -- topic of research in human cognition. The creative use of conceptual knowledge is, for example, a core element of scientific investigation, wherein the generation of new ideas is critical to designing experiments and explaining observations. However, due to its complexity, little is presently known about the cognitive processes underlying the generation of new concepts.

Towards the aim of providing a formal account of the cognitive mechanisms involved in creativity and innovation, we explore a simplified form of creative cognition that can be studied using standard behavioral methodology and is amenable to formal modeling: category generation. In a category generation task, participants are given some background knowledge about a domain (a cover story and/or learning one or more categories in the domain) and then are asked to generate another category in the domain \citep{jern2013probabilistic,ward1994structured}. Unlike creativity and innovation, categorization is well studied from a breadth of perspectives, including behavioral, neural, comparative, and computational perspectives \citep{kurtz2015human,mack2013,margolis2015,pothoswills2011}. By analyzing a creative task using formal tools from the categorization literature,

Previous work has focused on one key aspect of category generation: People tend to create new categories that have similar {\em statistical regularities} to previously known categories \citep{jern2013probabilistic,ward1994structured}. Although this is an important characteristic of generating new categories, it cannot be the only one. Taken to the extreme, the best "new" category in terms of having the same statistical regularities to other categories would be one that is identical to a prototypical previously observed category (and thus, not new at all).

In this article, we propose a new constraint that guides category generation: "being different" or contrasting from other categories in the relevant domain. To some extent, new categories being different in some sense from previously observed ones has been implicitly assumed in much previous work. But "being different" is vague as there are many ways to be different from a previously observed category. Building on the largely successful exemplar modeling framework \citep{medin1978context,nosofsky1984choice}, we propose a novel exemplar model of category generation, {\em Producing Alike and Contrasting Knowledge using Exemplar Representations} (PACKER), that formalizes the constraint of "being different" from previous categories while similar to other exemplars generated so far in the new category.

The outline of the article is as follows. First we describe the results of previous behavioral experiments investigating category generation. Next we review previous computational models of category generation. Then we propose our computational model, which produces new categories that are different from previous categories in the domain. Next we present two experiments examining how people generate new categories, one of which demonstrates that participants produce different categories when given one of two categories that do not differ in distributional information. We qualitatively and quantitatively analyze the performance of each model in capturing human category generation. We conclude with a discussion of the implications of our results for categorization and creative cognition, and directions for future work.

\section{Prior work}

Much of what we know about concept generation comes from the foundational literature on creative cognition. In a classic series of reports, Ward \& colleagues \citep{ward1995s,ward1994structured,marsh1999inadvertent,ward2002role,smith1993constraining} established that category generation is highly constrained by prior knowledge: Generated categories tend to consist of features observed in known categories, and they tend to exhibit the distributional properties as found in known categories. In a classic study, \cite{ward1994structured} asked participants to generate new species of alien animals by drawing and describing members of the species. People tended to generate species with the same features as on Earth (e.g., eyes, legs, wings), and possessing the same feature correlations as on Earth (e.g., feathers co-occur with wings). Likewise, aliens drawn from the same species tended to share more features with one another compared to members of opposite species.

@@

\jlanote{Yeah, Here's how I would this section should be organized. Top is general behavioral findings/phenomena. Then subsections for each theory, which first is described in broad theoretical terms and then how that is formalized into a model}

\cite{jern2013probabilistic} additionally tested a "copy-and-tweak" model that broadly resembles the earlier "Path of Least Resistance" account: the core proposal is that participants generate new items by copying stored examples from memory and tweaking them to generate something new. The copy-and-tweak model differs from the path of least resistance account in that it notably omits the hierarchical organization of categories, as well as selectivity in which features are changed. Instead, their copy-and-tweak model corresponds to a direct exemplar-similarity approach \citep[e.g.,][]{nosofsky1984choice}: participants are thought to generate new items according to their similarity to known members of the target category. This model provided a poor account of the observed data: indeed, the experiments devised by \citeauthor{jern2013probabilistic} were specifically designed to challenge this model. However, its application is notable as a first step toward explaining category generation using well-known formal approaches from the categorization literature.

@@

It is worth noting that the literature reviewed above provides a somewhat limited view of category generation: The main concern of the published experiments has been on the distributional correspondences between learned and generated categories, and as a result most of the modeling efforts have been towards explaining those effects. In this paper, we investigate another important constraint on the creative use of conceptual knowledge: category contrast. In order to generate a novel concept, individuals must produce something that is in some capacity \textit{different} from what they already know. Thus, in a trivial sense, contrast can be viewed as a fundamental constraint on creative generation: new concepts must firstly be different from existing ones.

Although it is evident that people are \textit{capable} of creating new concepts and categories, it is not entirely clear how new concepts are systematically made different from what is already known. The hierarchical sampling model developed by \cite{jern2013probabilistic}, for example, does not provide any strong claims about how generated categories should contrast with learned ones: the model only proposes that generated categories are sampled from the same underlying domain distribution as observed categories, and will thus share a common distributional structure. The authors do not make predictions about the \textit{location} of the category within the domain (the perceptual instantiation of category members). Indeed, under a strict interpretation of the hierarchical Bayesian approach presented by \citeauthor{jern2013probabilistic}, the most probable new category to be generated is located in {\em exactly} the same location as the given category. However, this is not simply an issue with their model -- Hierarchical Bayesian models assume the same underlying distribution generates all of the categories and thus, any differences between categories is due to {\em noise} and should not be {\em systematic}. The best a a hierarchical Bayesian model can do at capturing contrast is to assume that the new category is placed uniformly at random over stimulus space (this was the case in the simulations conducted by \citeauthor{jern2013probabilistic}).

@@

\jlanote{Don't worry about that for now. I'll do a pass on that sort of language later. Lets get the main stuff in here in the right places first.}

@@

\nbcnote{Maybe mention something about path of least resistance prediction that contrast is on non-characteristic dimensions? }
\jlanote{If we can make a strong case via verbal argument, than yeah. I think we decided not to model it, in which case we shouldn't go into detail on it.}

\section{Formalizing category generation}
\jlanote{this should come before the studies as formalizations of the different theories (some details may be delegated to Appendices). I like trying to integrate them with the verbal description of the theories rather than having a separate computational modeling section. But the latter works for sure and I'm not sure if the former will flow well...}
\subsection{PACKER}
The PACKER model is an extension of the influential Generalized Context Model of category learning \citep[GCM;][]{nosofsky1984choice,nosofsky1986attention}. The GCM simulates categorization under the assumption that learners represent categories as a collection of exemplars, corresponding to the labeled stimuli they have observed. The exemplars are encoded within a $k$-dimensional psychological space, and categorization is based on the amount of similarity between the item to be categorized and the stored exemplars. Similarity between two examples, $s\left(x_i, x_j\right)$, is computed as an inverse exponential function of distance \citep[following][]{shepard1957stimulus,shepard1987toward}:

\jlanote{K, though if I remember, it's just conjugacy, in which case we can just cite a paper that has IW-W conjugacy.}

\subsection{Synopsis and Prognosis}

Research on the creative generation of novel concepts has focused on the finding that generated categories tend to possess distributional commonalities with known categories. However, a fundamental goal of concept generation is to create something \textit{new} (i.e., different from what is already known). The manner in which generated categories differ from known ones is, nonetheless, poorly understood: existing formal approaches do not make strong predictions about how creatively generated concepts should systematically differ from existing ones.

@@

\jlanote{Is it possible to get the main individual differences we find in this study from different parameterizations of PACKER? If so, we can motivate it from there. Alternatively, we may want to switch the order of the two studies and motivate the experiment from the next one as a . I like switching the order of the experiments less because I feel like Experiment 2 nails contrast down and Experiment 1 is more showing how it plays a role in more complex category types}

@@
